# MFRE R Prep Section

All of you should have gone through the Python prep section by now and should be ready to start the R prep section. This section is designed to help you get familiar with the R programming language.

Many of the concepts you will have learned for Python will be applicable to R, but there are some key differences in syntax and functionality.

## R markdown

While the most basic way to run R code is to write in a `.r` file and use the R console, we will be using R markdown files to run our code. R markdown files are a great way to combine code, output, and text in a single document.

Rmd or R markdown files can be edited using your IDE or integrated development environment of your choice, but it can also be edited using a simple text editor like Notepad. The advantage of using Rmd over a simple R script is that you can include text, code, and output in a single document and have it be rendered or knitted into a variety of formats like HTML, PDF, or Word.

Similar to Jupyter notebooks, R markdown files are divided into cells. Each cell can contain either text or code. You can run the code in a cell by pressing the run button or by using the keyboard shortcut `Ctrl + Enter`.

Here is a code example:

```{r}
first_value <- 5
second_value <- 10
result <- first_value + second_value
print(result)
```

Here we see the first difference between the Python and R syntax. In R, we use the `<-` operator to assign values to variables. The code above assigns the value 5 to the variable `x`, the value 10 to the variable `y`, and then adds `x` and `y` together and assigns the result to the variable `z`. The code then prints the value of `z`.

For those who are familiar with other languages, R is syntactically similar to C and C++.

# Introduction to R

R is a programming language and free software environment for statistical computing and graphics supported by the R Foundation for Statistical Computing. The R language is widely used among statisticians and data miners for developing statistical software and data analysis. While you will rarely use R for general programming, it is a powerful tool for data analysis and visualization due in part to the litany of packages available for R.

## Object-Oriented Programming vs Functional Programming

R is a functional programming language, which means that it is designed to evaluate mathematical functions and avoid changing state and mutable data. This is in contrast to object-oriented programming languages like Python, which gives you the option of doing functional programming but are primarily designed to work with objects and classes.

For our purposes, this means that unlike Python where we can access methods and properties of objects, in R we will be working with functions and data frames.

Here's an example of this difference in action:

```python
# Python code
array_1 = [1, 2, 3, 4, 5]
array_1.append(6) # This dot notation is used to access the append method of the list object
print(array)
```

```{r}
# R code
array_1 <- c(1, 2, 3, 4, 5)
array_1 <- c(array_1, 6)
print(array_1)

# Notice that comments in R are denoted by a hash symbol as well
```

**Learning Objectives**

We will go through the following topics in this section:

1. Data types
2. Data structures 
3. Basic operations 
4. Control flow 
5. Functions 
6. Variable naming conventions 
7. The joys of pseudocode and how to write it
8. Rubber duck debugging
9. R Resources

## Data Types

R has a number of data types, some of which are similar to Python and some of which are unique to R. Here are some of the most common data types in R:

1. **Numeric** - This is the most common data type in R and is used to store numbers. Numeric data types can be integers (whole numbers) or floats (numbers with a decimal).
2. **Character** - This data type is used to store strings.
3. **Logical** - This data type is used to store boolean values, i.e. `TRUE` or `FALSE`.
4. **Integer** - This data type is used to store whole numbers.
5. **Complex** - This data type is used to store complex numbers.
6. **Raw** - This data type is used to store raw bytes.

Here are some examples of how to define variables of different data types in R:

```{r}
# Numeric
n_value <- 5
print(n_value)

# Character
c_value <- "Hello, World!"
print(c_value)

# Logical
l_value <- TRUE
print(l_value)

# Integer
i_value <- as.integer(5)
print(i_value)

# Complex
c_value <- 5 + 3i
print(c_value)

# Raw
r_value <- charToRaw("Hello, World!")
print(r_value)
```

Mixing data types in R can lead to unexpected results, so it's important to be aware of the data types you are working with. Here are some examples of how mixing data types can lead to unexpected results:


```{r}
# Adding a numeric and a character
n_value <- 5
c_value <- "10"
tryCatch(
    {
        result <- n_value + c_value
        print(result)
    },
    error = function(e) {
        print(e)
    }
)

# Adding a numeric and a logical
n_value <- 5
l_value <- TRUE
result <- n_value + l_value
print(result)
```

## Data Structures

R has a number of data structures that are used to store and manipulate data. Some of the most common data structures in R are:

1. **Vectors** - Vectors are one-dimensional arrays that can store numeric, character, or logical data. Vectors can only store data of the same type.
2. **Matrices** - Matrices are two-dimensional arrays that can store numeric, character, or logical data. Matrices can only store data of the same type.
3. **Lists** - Lists are one-dimensional arrays that can store data of different types.
4. **Data Frames** - Data frames are two-dimensional arrays that can store data of different types. Data frames are similar to matrices, but are more flexible and can store data of different types.

Here are some examples of how to define and work with vectors, matrices, lists, and data frames in R:

```{r}
# Vectors
vec <- c(1, 2, 3, 4, 5)
print(vec)

# Matrices
mtx <- matrix(1:9, nrow = 3, ncol = 3)
print(mtx)

# Lists
li <- list(1, "Hello, World!", TRUE)
print(li)

# Data Frames
# This is the closest analogue to a pandas dataframe in Python
df <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))
print(df)
```

### Basic Operations

R supports all the basic arithmetic operations that you would expect. Here are some examples:

```{r}
first_value <- 5
second_value <- 10
result <- first_value + second_value
print(result)

result <- first_value - second_value
print(result)

result <- first_value * second_value
print(result)

result <- first_value / second_value
print(result)

result <- first_value %% second_value
print(result)

result <- first_value ^ second_value
print(result)
```

**Side Note** : Here's how you print a formatted string in R:

```{r}
first_value <- 5
second_value <- 10
result <- first_value + second_value
print(sprintf("The sum of %d and %d is %d", first_value, second_value, result))
```

## Control Flow

Control flow in R is similar to control flow in other programming languages. R supports if-else statements, for loops, and while loops. Here are some examples of how to use control flow in R:

```{r}
# If-else statements
value <- 5
if (value > 0) {
    print("value is positive")
} else {
    print("value is negative")
}

# For loops
for (i in 1:5) {
    print(i)
}

# While loops
i <- 1
while (i <= 5) {
    print(i)
    i <- i + 1
}

# Boolean operators
first_value <- 5
second_value <- 10
if (first_value > 0 & second_value > 0) {
    print("Both x and y are positive")
}

if (first_value > 0 | second_value > 0) {
    print("Either x or y is positive")
}

if (!first_value > 0) {
    print("x is not positive")
}

# In statements
hello_world <- "Hello, World!"
if ("H" %in% hello_world) {
    print("H is in x")
}
```

**Note**: R uses curly braces to denote code blocks, unlike Python which uses indentation. 

## Functions

Functions are key to writing readable, modular, and efficient code. In R, you can define functions using the `function` keyword. Here's an example of how to define and use a function in R:

```{r}
# Define a function
add <- function(first_value, second_value) {
    result <- first_value + second_value
    return(result)
}

# Call the function
result <- add(5, 10)
print(result)
```

Unlike Python, the def prefix is not used to define functions in R. Instead, you use the `function` keyword followed by the function name and the arguments in parentheses. The function body is enclosed in curly braces, and the return value is specified using the `return` keyword.

It is also possible to call other functions from within a function. Here's an example:

```{r}
# Define a function
add <- function(first_value, secodn_value) {
    result <- first_value + second_value
    return(result)
}

# Define another function
square <- function(value) {
    result <- add(value, value)
    return(result)
}

# Call the function
result <- square(5)
print(result)
```

**Note**: You have to define a function before you can call it. If you try to call a function that hasn't been defined yet, you will get an error.

Recursive functions are functions that call themselves in their own definition. They are used to solve problems that can be broken down into smaller, similar problems. The classic examples of recursive functions are factorial and Fibonacci sequence calculations.

Here's an example of a recursive function in R that calculates the factorial of a number:

```{r}
factorial_recursive <- function(n) {
    # Base case: factorial of 0 is 1
    if (n == 0) {
        return(1)
    } else {
        # Recursive case: n * factorial of (n-1)
        return(n * factorial_recursive(n - 1))
    }
}

# Call the function
print(factorial_recursive(5)) # Outputs: 120
```

In this example, `factorial_recursive()` calls itself to calculate the factorial of n. The function has a base case that ends the recursion `(if (n == 0))`, and a recursive case that calls the function again with a smaller argument `(return(n * factorial_recursive(n - 1)))`.

Recursive functions are useful when you're dealing with problems that have a recursive structure, such as traversing tree-like data structures, solving puzzles like the Towers of Hanoi, or performing certain mathematical calculations. However, they can be harder to understand and debug than iterative solutions, and they can cause a stack overflow if the recursion goes too deep. Therefore, they should be used judiciously.

## Variable Naming Conventions

In R, variable names can contain letters, numbers, and underscores, but they cannot start with a number. Variable names are case-sensitive, so `x`, `X`, and `X1` are all different variables. It is a good practice to use descriptive variable names that reflect the purpose of the variable. Here are some examples of valid and invalid variable names in R:

```{r, eval=FALSE}
# Valid variable names
first_value <- 5
y <- "Hello, World!"
z_1 <- TRUE
my_variable <- 10

# Invalid variable names
1x <- 5
hello world <- "Hello, World!"
my-variable <- 10
```

It is also a good practice to follow a consistent naming convention for variables. Some common naming conventions in R include:

1. **CamelCase**: The first letter of each word is capitalized except for the first word, e.g., `myVariableName`.
2. **snake_case**: Words are separated by underscores, e.g., `my_variable_name`.
3. **PascalCase**: The first letter of each word is capitalized, e.g., `MyVariableName`.

Choose a naming convention that you are comfortable with and stick to it throughout your code to make it more readable and maintainable. This is especially important when working on larger projects or collaborating with other developers.

**Tip** : Use descriptive variable names that reflect the purpose of the variable, e.g., `total_sum` instead of `x`. Try to also keep your variable names short and concise which will help when trying to type them out later on, but not so short that they become cryptic.

## Pseudocode

Pseudocode is an informal high-level description of the operating principle of a computer program or other algorithm. It uses the structural conventions of a programming language, but is intended for human reading rather than machine reading. Pseudocode is often used to plan and describe algorithms before writing the actual code.

This can be a useful tool when you are trying to figure out how to solve a problem or implement an algorithm. Writing pseudocode can help you break down the problem into smaller steps and think through the logic of your solution before you start coding. Pseudocode is not syntax-specific, so you can write it in plain English or any other language you are comfortable with and even implement the same code in different languages.

Pseudocode is not a formal language, so there are no strict rules for writing it. However, there are some common conventions that can help make your pseudocode more readable and effective:

1. **Use descriptive variable names**: Use variable names that describe the purpose of the variable, e.g., `total_sum` instead of `x`.
2. **Indentation**: Use indentation to show the structure of your code, e.g., indent the body of a loop or conditional statement.
3. **Comments**: Use comments to explain the purpose of each step or section of your pseudocode.
4. **Avoid technical jargon**: Write your pseudocode in plain language that is easy to understand, even for non-programmers.
5. **Break down complex steps**: Break down complex steps into smaller, more manageable steps to make your pseudocode easier to follow.

This will also form the basis of good descriptive comments in your code. 

Here's an example of pseudocode for finding the sum of the first n natural numbers:

```
# Pseudocode for finding the sum of the first n natural numbers
# Input: n (a positive integer)
# Output: sum (the sum of the first n natural numbers)

# Initialize sum to 0
sum = 0

# Loop from 1 to n
for i = 1 to n
    # Add i to sum
    sum = sum + i

# Output the sum
return sum
```

This pseudocode describes the algorithm for finding the sum of the first n natural numbers. It initializes a variable `sum` to 0, then loops from 1 to n, adding each number to the sum. Finally, it returns the sum as the output.

Writing pseudocode can help you plan and organize your code before you start writing it in a specific programming language. It can also be a useful tool for communicating your ideas to others and documenting your thought process as you work through a problem.

## Rubber Duck Debugging

Rubber duck debugging is a method of debugging code by explaining it to an inanimate object, such as a rubber duck. The idea is that by explaining your code step by step, you can identify and fix errors more easily. This technique is based on the principle that explaining a problem out loud can help you understand it better and find a solution.

Here's how rubber duck debugging works:

1. **Get a rubber duck**: Find an inanimate object, such as a rubber duck, to act as your debugging partner.
2. **Explain your code**: Go through your code line by line, explaining each step to the rubber duck in plain language.
3. **Identify errors**: As you explain your code, you may notice errors or inconsistencies that you hadn't seen before.
4. **Fix the errors**: Once you've identified the errors, make the necessary changes to your code.
5. **Repeat as needed**: If you encounter more errors, continue explaining your code to the rubber duck until you've resolved all the issues.

Rubber duck debugging can be a helpful technique when you're stuck on a problem or having trouble finding a bug in your code. By verbalizing your thought process and explaining your code to an external observer, you may gain new insights and uncover hidden issues that you hadn't noticed before.

**Additional tips for debugging**: 

1. **Break the problem down**: If you're having trouble debugging a complex piece of code, try breaking it down into smaller parts and testing each part separately.
2. **Use print statements**: Insert print statements in your code to track the values of variables and see how the code is executing.
3. **Check your assumptions**: Make sure you're not making any incorrect assumptions about how your code should work.
4. **Take a break**: If you're feeling frustrated or stuck, take a break and come back to the problem with fresh eyes.
5. **Read the error messages**: Pay attention to any error messages or warnings that your code produces, as they can provide valuable information about what's going wrong.

## R Resources

R is a powerful and versatile programming language that is widely used for data analysis, statistical modeling, and visualization. If you're new to R, there are many resources available to help you learn the language and become proficient in using it. Here are some recommended resources for learning R:

1. **R Documentation**: The official R documentation is a comprehensive resource that covers all aspects of the R language, including functions, packages, and data structures. You can access the R documentation [online](https://cran.r-project.org/manuals.html) or within the R console by typing `?function_name` to get help on a specific function.

2. **R for Data Science**: This book by Hadley Wickham and Garrett Grolemund is a popular resource for learning R for data analysis. It covers the basics of R programming, data manipulation, visualization, and modeling using the `tidyverse` package.

3. **Coursera**: Coursera offers a number of courses on R programming, data analysis, and statistics. Some popular courses include "R Programming" and "Data Science and Machine Learning Bootcamp with R".

4. **DataCamp**: DataCamp is an online platform that offers interactive courses on R programming, data analysis, and machine learning. It is a great way to learn R through hands-on exercises and projects.

5. **Stack Overflow**: Stack Overflow is a popular question and answer site for programming-related questions. If you run into problems or have questions about R programming, you can search for answers on Stack Overflow or ask your own question.

6. **RStudio Community**: The RStudio Community is a forum for R users to ask questions, share tips and tricks, and connect with other R programmers. It's a great place to get help and advice on R programming.

7. **GitHub**: GitHub is a code hosting platform where you can find R packages, projects, and resources. You can explore R repositories on GitHub to find useful code snippets, packages, and tutorials.

8. **R Bloggers**: R Bloggers is a blog aggregator that collects posts from R bloggers around the world. It's a great way to stay up to date on the latest news, tutorials, and tips related to R programming.

These are just a few of the many resources available for learning R. Whether you're new to R or an experienced user looking to expand your skills, there are plenty of opportunities to learn and grow as an R programmer.

# Final Notes About R and Rmd

Unlike Jupyter Notebook, Rmd files are not interactive. This means that you can't run a cell and then run another cell that depends on the output of the first cell. Instead, you have to run the entire Rmd file to see the output of all the cells. This can be a bit cumbersome, but it's a small price to pay for the flexibility and power of Rmd files.

Another important thing to note is that Rmd files are rendered or knitted into different formats using an R package called `knitr`. This package allows you to convert your Rmd file into HTML, PDF, or Word format. You can specify the output format in the YAML header at the top of your Rmd file. Here's an example of a YAML header that specifies the output format as HTML:

```{r, eval=FALSE}
---
title: "My R Markdown Document"
output: html_document
---
```

# Assignment

1. Construct a 2 x 3 matrix which takes in values ranging from 1 to 6 in decending order row wise.
```{r}
# Your code here
```

2. Write a function that prints a pyramid using the `*` character. The function should take in a positive integer as an argument, and print a pyramid with that many levels. 

```{r}
# Your code here
```

3. The following function is supposed to return the cumulative sum of numbers from 1 to n. However, there is a bug in the code. Find and fix the bug.

```{r}
# Your code here
cum_sum <- function(n) {
    sum <- 0
    for (i in 1:n) {
        sum <- i
    }
    return(sum)
}

print(cum_sum(5))
```

4. You are given a 16 digit credit card number as a string. Write a function that takes in the credit card number as an argument and returns `TRUE` if the credit card number is valid and `FALSE` if it is not. A credit card number is valid if the sum of the digits is divisible by 10.

```{r}
# Your code here
string <- "1234567890123456"
```

5. You are given a 3 x 3 dataframe with the rows representing cities and the columns representing dates. Identify if any of the cities have a temperature above 30 degrees on any of the dates and add a new column to the dataframe with the value `TRUE` if the city has a temperature above 30 degrees and `FALSE` if it does not.

```{r}
# Your code here
df <- data.frame(
    city = c("City A", "City B", "City C"),
    date_1 = c(25, 30, 35),
    date_2 = c(28, 29, 30),
    date_3 = c(30, 32, 33)
)
```